---
title: "Challenge 7: Incorporating Multiple Inputs"
subtitle: "Functions + Fish"
author: "Matteo Saisi"
format: html
editor: source
embed-resources: true
code-fold: true
code-tools: true
message: false
warning: false
---

```{r}
#| label: setup

library(tidyverse)
library(here)

BlackfootFish <- read.csv(here::here("data", "BlackfootFish.csv"))

```

## Condition Index

A frequently used measurement for fish health is a condition index [(Wikipedia article)](https://en.wikipedia.org/wiki/Condition_index_in_fish). The following
simple equation can be used to calculate the approximate condition index of a
fish:

$$\text{condition index} = \frac{weight}{length^3} \times 100$$

**1. There are specific units required for the calculation of a condition index -- length must be in centimeters and weight must be in grams. The weight data for the Blackfoot River fish were collected in grams; the length data were collected in millimeters. Transform the length data to the correct units.**

```{r}
#| label: transform-data-to-have-correct-units

BlackfootFish <- BlackfootFish |>
  mutate(length = length / 10)

```

**2. Collecting data of this sort can be very messy! Write a function that will replace unlikely `length` and `weight` measurements with `NA`. Your function should accept at least three inputs:**

+ **a vector of measurements,**
+ **the minimum reasonable value,**
+ **the maximum reasonable value.**

**If a value falls outside these bounds, you should replace it with an `NA`.**

::: callout-tip
If you are struggling with the structure of your function, I would suggest reading the [Mutating Function](https://r4ds.hadley.nz/functions.html#mutate-functions) from R4DS.
:::

```{r}
#| label: function-to-change-impossible-measurements-to-NA

unreasonable_values <- function(values, min_measurement, max_measurement){
  case_when(
    values < min_measurement ~ NA,
    values > max_measurement ~ NA,
    .default = values
  )
}

```

**3. After consulting the [Montana Record Table](https://fwp.mt.gov/fish/anglingData/records) for the four species of trout included in these data, I have conjectured that it is unlikely to have measurements for fish below 5 cm and above 80 cm in length or below 10 g and above 4,000 g in weight. Use your function to modify the `length` and `weight` columns of the `fish` dataset based on my cutoffs.**

```{r}
#| label: modify-impossible-values-of-length-weight

BlackfootFish <- BlackfootFish |>
  mutate(length = unreasonable_values(length, 5, 80),
         weight = unreasonable_values(weight, 10, 4000))

```

**3. Write a function to calculate the condition index of a fish, given inputs of weight and length.**

::: callout-warning
Consider whether you will write a function that works with vectors (accepting
vectors as inputs) or a function that works with data frames (accepting a 
dataset and variable names as inputs)!
:::

```{r}
#| label: condition-function

condition_index <- function(df, var_1, var_2){
  df |>
    mutate(c_index = ({{var_1}} / ({{var_2}} ** 3)) * 100)
}

```

**4. Make a thoughtful visualization of how fish conditions have varied over the duration of this study.**

::: callout-warning
### To be thoughtful...

... you need to use at least a third variable!
:::

```{r}
#| label: condition-indices-over-time

BlackfootFish |>
  condition_index(weight, length) |>
  ggplot(mapping = aes(x = year, y = c_index, color = section)) +
  geom_smooth(na.rm = TRUE) +
  facet_wrap(facets = "species") +
  scale_color_manual(values = c("black", "deepskyblue2"),
                     aesthetics = "color") +
  labs(x = "Time (Years)",
       y = "",
       subtitle = "Condition Index",
       title = "The Condition Index of Weight and Length over Time",
       color = "Section")

```
